Prediction of 2024 US election ...

### 2024-11-04 paper/paper.qmd
colin code update6

library(tidyverse)
library(here)
library(tibble)
library(knitr)
library(maps)
library(mapproj)
light_blue <- "#ADD8E6"
dark_blue <- "#6688CC"
light_red <- "#FF7777"
dark_red <- "#B22222"
purple <- "#9370DB"
```

@@

For the data we used in this analysis about the polling result for Kamala Harris and Donalad Trump in 2024 USA president election. \
- **response variable:** pct(pct: The percentage of the vote or support that the candidate received in the poll) \
- **numeric predictor:**\
 sample size(sample_size: The total number of respondents participating in the poll) \
 timegap(the time gap between the poll start date and the real election date i.e timegap = real US election date - poll start date) \
 pollscore(A numeric value representing the score or reliability of the pollster in question) \
- **categorical predictor**
state(The U.S. state where the poll was conducted or focused) \
methodology(The method used to conduct the poll)\

@@

## Clean Data
The data cleaning process involves several steps to ensure the quality and relevance of the polling data. First, we filter the dataset to retain only poll results with a numeric grade of 2.7 or higher, indicating that the polls are considered reliable. Next, we address missing values in the state attribute: polls with NA in the state column are considered national polls.

@@

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Sample of cleaned US election data 
#| echo: false
#| warning: false
cleaned_data = read_csv(here("data/02-analysis_data/cleaned_data.csv"))

cleaned_data |>
  select(pct, sample_size, pollscore, days_taken_from_election, state, methodology, candidate_name) |>
  head(6) |>
  kable(
    col.names = c("pct", "sample_size", "pollscore", "days_taken_from_election","state", "methodology", "Candidate Name"),
    booktabs = TRUE
    ) 
```

### States included in anaylsis

After the data cleaning process, 21 states had no polling data. A table showing the number of polls for each state, including those without any polls, is provided in @tbl-state-polls-count.

This absence of polling data is not a significant concern due to the structure of the United States Electoral College (explained in detail in the Appendix). The states lacking polling data have consistently followed historical voting patterns, so predicting the winning candidate in those states is unnecessary

@@

```{r}
#| label: fig-avg-pct-state
#| fig-cap: Historical State Voting Trends Are Maintined in 2024
#| echo: false
#| warning: false
#| 
cleaned_data <- cleaned_data %>% mutate(candidate_name =
  if_else(!(candidate_name %in% c("Kamala Harris", "Donald Trump")), "Other", candidate_name))

cleaned_data %>%
  filter(candidate_name %in% c("Kamala Harris", "Donald Trump")) %>%
  group_by(state, candidate_name) %>%
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop") %>%
  ggplot(mapping = aes(x = avg_pct, y = state, fill = candidate_name)) + 
  geom_col(position = "stack") +
  theme_minimal() +
  labs(
    x = "Average Polled Support (%)", 
    y = "State",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = light_blue, "Donald Trump" = light_red)) +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  theme(axis.text.y = element_text(size = 8))

```
In figure @fig-avg-pct-state, historically Democratic are polling for Kamala and historically Republican states are polling for Trump. Similarly, historically swing states also appear to be close, for instance Michigan (46.9% Trump, 47.5% Harris), Nevada (47.7% Trump, 47.9% Harris), and Pennsylvania (46.9% Trump, 48.2% Harris) all are close to an even split.

```{r}
#| label: fig-avg-pct-methodology
#| fig-cap: Polls with more Methodologies have Less Variability.
#| echo: false
#| warning: false
#| 
cleaned_data <- cleaned_data %>% mutate(candidate_name =
  if_else(!(candidate_name %in% c("Kamala Harris", "Donald Trump")), "Other", candidate_name))

cleaned_data %>%
  filter(candidate_name %in% c("Kamala Harris", "Donald Trump")) %>%
  filter(!is.na(methodology)) %>% 
  ggplot(mapping = aes(x = methodology, y = pct)) + 
  geom_boxplot(width = 0.6, position = position_dodge(width = 0.8), outlier.shape = NA) +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "Methodology",
    y = "Polled Support (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = light_blue, "Donald Trump" = light_red)) +
  theme(legend.position = "bottom") + 
  theme(axis.text.y = element_text(size = 8))

```
Figure @fig-avg-pct-methodology shows that polls utilizing multiple communication methods to reach voters tend to have lower interquartile ranges (IQRs) in their boxplots compared to those that rely on only one or two communication methods.

## Relationship Between Variables

```{r}
#| label: fig-pollscore-samplesize
#| fig-cap: More Reliable Pollsters Have Larger Sample Sizes in their Polls
#| echo: false
#| warning: false

# removing outliers and na's
graph_data <- cleaned_data %>% 
  filter(!is.na(pollscore) & !is.na(sample_size))

outlier_top <- (quantile(graph_data$sample_size, 0.75) + 
                  IQR(graph_data$sample_size)*1.5)

outlier_bottom <- (quantile(graph_data$sample_size, 0.25) - 
                  IQR(graph_data$sample_size)*1.5)

graph_data <- cleaned_data %>%
  filter(sample_size < outlier_top & sample_size > outlier_bottom)

graph_data %>% filter(!is.na(pollscore) & !is.na(sample_size)) %>% 
  ggplot(aes(x = pollscore, y = sample_size)) +
  geom_point(alpha = 0.33) + geom_smooth(method = "lm", se = FALSE, color = light_red) +
  labs(x = "Poll Score", y = "Poll Sample Size")

# this model is really here just for explain, this realtionship should not
# be treated as lienar
# cor(graph_data$sample_size, graph_data$pollscore)
# lm_score_sample <- lm(sample_size ~ pollscore, data = graph_data)
# summary(lm_score_sample)
```
Figure @fig-pollscore-samplesize illustrates a weak positive correlation between a pollster's pollscore and the sample size of their poll. The figure also shows that most polls have a sample size around 800 to 1200 participants. It is also important to note that a few polls with exceptionally large sample sizes were excluded from the graph due to their status as clear outliers.

```{r}
#| label: fig-pollsamplesize-canidate-pct
#| fig-cap: The Sample Size of a Poll does not impact a Candidate's Voting Percentage
#| echo: false
#| warning: false

cleaned_data %>% 
  filter(!is.na(sample_size) & !is.na(pct)) %>% 
  ggplot(aes(x = sample_size, y = pct)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = "lm", se = FALSE, color = light_red) +
  labs(x = "Sample Size of the Poll",
       y = "Candidate Percentage") + facet_wrap(~candidate_name)

```
Based on @fig-pollsamplesize-canidate-pct, there does not seem to be a relationship between the sample size of a poll the percentage of a candidate.

```{r}
#| label: fig-pollscore-canidate-pct
#| fig-cap: More Reliable Pollsters Score Higher For Trump
#| echo: false
#| warning: false

cleaned_data %>% 
  filter(!is.na(pollscore) & !is.na(pct)) %>% 
  ggplot(aes(x = pollscore, y = pct)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = "lm", se = FALSE, color = light_red) +
  labs(x = "Pollscore",
       y = "Candidate Percentage") + facet_wrap(~candidate_name)

```
Figure @fig-pollscore-candidate-pct depicts the relationship between a pollster's poll score and a candidate's percentage. For Donald Trump, a negative correlation is observed, indicating that pollsters with lower poll scores tend to assign him a higher percentage of support. Conversely, Kamala Harris shows the opposite trend: as the poll score decreases, her percentage tends to rise. This suggests that more reliable polls, characterized by lower poll scores, report higher support for Trump compared to less reliable pollsters.

Based on @tbl-state-candidate-support, state might have to be a removed variable.

```{r}
#| label: fig-polldate-canidate-pct
#| fig-cap: Both Canidate Percentages Are Gaining Support Closer to the Election
#| echo: false
#| warning: false

cleaned_data %>% 
  filter(!is.na(days_taken_from_election) & !is.na(pct)) %>% 
  ggplot(aes(x = days_taken_from_election, y = pct)) +
  geom_point(alpha = 0.4) + 
  geom_smooth(method = "lm", se = FALSE, color = light_red) +
  labs(x = "Timing of Polls Relative to Election Day (Days Prior to the Election)",
       y = "Candidate Percentage") + facet_wrap(~candidate_name)

```
Based on @fig-polldate-canidate-pct, Trump and Harris are getting more votes in polls that are done closer to the election. This is a result of non-major candidate support rapidly decreasing. Specifically, Trump support is increasing at a faster rate than Harris. However, the polls in general show a slight lead for Harris throughout the last 100 days.

The rapid decline in third party support could be due to Robert F. Kennedy dropping out of the race (this sentences should probably be in results or discussion section).

\newpage
# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

@@

@@

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

@@

# Overview of American Electon System

## Brief Description of American Federal Goverment

The American federal or national government is split into three branches, executive., legislative, and judical The executive branch includes the president and the military. The legislative branch include two subgroups, the House of Representatives and the Senate. These two subgroups create laws. Every state has two Senators and one Representative per approximately 750,000 people. The judicial branch is court system.

## What is the Electoral College {#sec-electoral-college}

The Electoral College is the system used in the United States to elect the president and vice president. Instead of a direct popular vote, each state is allocated a certain number of electors based on its representation in Congress (the total of its Senators and Representatives). When voters cast their ballots, they are actually voting for a slate of electors pledged to a candidate. The candidate who receives a majority of electoral votes (270 out of 538) wins the presidency. This system means that winning the popular vote in a state generally results in winning all of that state's electoral votes. The only exception are the states of Maine and Nebraska, who award electoral votes by congressional district, with two additional votes given to the statewide winner.

The Electoral College results in some states being unnecessary to campaign in, as their strong historical voting pasterns towards either Democrats or Republicans make them unlikely to change, regardless of campaign efforts. Therefore, for statisticians, polling information from these states may not be that useful when trying to predict the outcome of an election. On the other hand, states that can vote either Democratic or Republican (swing states) are immensely important when predicting an election. As a result, campaigns spend hundreds of millions of dollars campaigning and understanding voters there.

```{r}
#| label: fig-currentstateofelection
#| fig-cap: 2024 U.S. Presidential Election State Forecast Map
#| echo: false
#| warning: false

# Define your colors for each state (replace these colors with your desired ones)
state_colors_legend <- c(
  "Republican" = dark_red,      # Dark Red
  "Lean Republican" = light_red, # Light Red
  "Toss-Up" = purple,         # Purple
  "Lean Democrat" = light_blue,   # Light Blue
  "Democrat" = dark_blue,        # Dark Blue
  "Split By District" = "gray"
)

state_colors <- c(
  alabama = dark_red,
  arizona = purple,
  arkansas = dark_red,
  california = dark_blue,
  colorado = dark_blue,
  connecticut = dark_blue,
  delaware = dark_blue,
  `district of columbia` = dark_blue,
  florida = light_red,
  georgia = purple,
  idaho = dark_red,
  illinois = dark_blue,
  indiana = dark_red,
  iowa = light_red,
  kansas = dark_red,
  kentucky = dark_red,
  louisiana = dark_red,
  maine = "gray",
  maryland = dark_blue,
  massachusetts = dark_blue,
  michigan = purple,
  minnesota = light_blue,
  mississippi = dark_red,
  missouri = dark_red,
  montana = dark_red,
  nebraska = "gray",
  nevada = purple,
  `new hampshire` = light_blue,
  `new jersey` = dark_blue,
  `new mexico` = light_blue,
  `new york` = dark_blue,
  `north carolina` = purple,
  `north dakota` = dark_red,
  ohio = light_red,
  oklahoma = dark_red,
  oregon = dark_blue,
  pennsylvania = purple,
  `rhode island` = dark_blue,
  `south carolina` = dark_red,
  `south dakota` = dark_red,
  tennessee = dark_red,
  texas = light_red,
  utah = dark_red,
  vermont = dark_blue,
  virginia = light_blue,
  washington = dark_blue,
  `west virginia` = dark_red,
  wisconsin = purple,
  wyoming = dark_red
)

# Load the states map data
states_map <- map_data(map = "state")

state_colors <- setNames(state_colors, tolower(names(state_colors)))

color_to_status <- setNames(names(state_colors_legend), state_colors_legend)

states_map <- states_map %>%
  mutate(status = color_to_status[state_colors[region]])

plotclr <- c(dark_red, light_blue, purple, gray, light_blue, dark_blue)
legend_description <- c(
  "Likely Republican",
  "Lean Republican",
  "Swing States",
  "Lean Democrat",
  "Likely Democrat",
  "Split by District"
)

ggplot(states_map, aes(x = long, y = lat, group = group, fill = status)) +
  geom_polygon(color = "white", size = 0.2) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_fill_manual(
    values = state_colors_legend,
    name = "Election Status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

```
The state statuses presented in this map are based on evaluations from @cnn, @foxnews, and @msnbc, which are generally agreed upon within the American political community. These organizations assessed historical voting patterns and recent polling data to derive their conclusions.

Notably, Nebraska and Maine are indicated in gray due to their delegates being split by district. In Nebraska, the state overall is projected to lean Republican, with the first and third districts also strongly favoring Republican candidates, while the second district leans Democratic. In Maine, the overall expectation is a Democratic leaning, consistent with its first district, though the second district leans Republican. Furthermore, Alaska and Hawaii are not in the map. Alaska is strongly favoring Republicans while Hawaii strongly favors democrats.

The seven generally agreed upon swing states are Pennsylvania, North Carolina, Georgia, Arizona, Nevada, Wisconsin, and Michigan with Texas, Florida, Nebraska District Two, Maine District 2, and Minnesota as the next closest races.

# States Poll Count

```{r}
#| label: tbl-state-polls-count
#| tbl-cap: Polls included in Analysis Per State
#| echo: false
#| warning: false

states <- c(state.name, "National")
polls_per_state_tbl <- tibble(state = states)

polls_per_state <- cleaned_data %>% group_by(state) %>% 
  summarise(polls_count = n_distinct(poll_id))

polls_per_state_tbl <- polls_per_state_tbl %>% 
  left_join(polls_per_state, by = "state") %>% 
  mutate(polls_count = if_else(!is.na(polls_count), polls_count, 0)) %>% 
  select(state, polls_count)

polls_per_state_tbl %>% kable(col.names = c("State", "Number of Polls In Analysis"))

num_states_missing <- count(polls_per_state_tbl %>% 
                              filter(polls_count == 0) %>% 
                              select(state))

# The Ideal Survey

## Objective

The research team has developed a survey and distribution methodology with a hypothetical budget of $100,000. This objective is to create a polling system that accurately predicts the 2024 United States Election.

## Sampling Frame

Given the monetary constraint, the team's first decision is to mostly include swing states and districts. These are Nevada, Arizona, Wisconsin, Michigan, Pennsylvania, North Carolina, and Georgia. Swing states are the primary focus of the poll because they disproportionately affect the outcome of the election based on the explanation in @sec-PLACEHOLDER. However, the poll will also include Texas, Florida, and Minnesota, Nebraska District 2, and Maine District 2, as they have the potential to one by either party but are not as likely to change as the previously mentioned swing states. Nevertheless, it is important to note that to save costs, the team will focus on polling true swing states compared to Texas, Florida and Minnesota.

## Survey Sending Process

>>>>

library(tidyverse)
library(palmerpenguins)
library(here)
library(tibble)
library(knitr)
library(brms)
library(rstanarm)
library(gridExtra)
library(Matrix)
library(rstan)
library(brms)

```{r}
#| echo: false
#| warning: false
cleaned_data = read_csv(here("data/02-analysis_data/cleaned_data.csv"))
```

@@

For the data we used in this analysis about the polling result for Kamala Harris and Donalad Trump in 2024 USA president election. \
- **response variable:** `pct`(pct: The percentage of the vote or support that the candidate received in the poll) \
- **numeric predictor:**\
 `sample size`(sample_size: The total number of respondents participating in the poll) \
 `timegap`(the time gap between the poll start date and the real election date i.e timegap = real US election date - poll start date) \
 `pollscore`(A numeric value representing the score or reliability of the pollster in question) \
- **categorical predictor**
`state`(The U.S. state where the poll was conducted or focused) \
`methodology`(The method used to conduct the poll)\

## Data Exploration
```{r}
#| label: tbl-explore-data
#| echo: false
#| warning: false
a <- ggplot(data=cleaned_data, aes(x=sample_size, y=pct)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) + 
  labs(x = 'sample size', y='pct', 
       title = 'pct vs sample size')

b <- ggplot(data=cleaned_data, aes(x=pollscore, y=pct)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) + 
  labs(x = 'pollscore', y='pct', 
       title = 'pct vs pollscore')

c <- ggplot(data=cleaned_data, aes(x=methodology, y=pct)) + 
  geom_boxplot() + 
  labs(x = 'methodology', y='pct', 
       title = 'pct vs methodology')

d <- ggplot(data=cleaned_data, aes(x=days_taken_from_election, y=pct)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) + 
  labs(x = 'max gust speed', y='pct', 
       title = 'pct vs days taken from election')

e <- ggplot(data=cleaned_data, aes(x=state, y=pct)) + 
  geom_boxplot() + 
  labs(x = 'state', y='pct', 
       title = 'pct vs state')

grid.arrange(a, b, c, d, e, nrow=2)
```

- pct vs sample size: This scatter plot shows the pct against the sample size, with a fitted trend line indicating a slight positive relationship. The data points are denser for lower sample sizes, suggesting that smaller sample sizes are more common in the dataset.

- pct vs pollscore: This scatter plot illustrates the pct against pollscore. The fitted trend line suggests a weak negative relationship between pollscore and pct. The points are scattered without a strong linear pattern.

- pct vs methodology: A boxplot comparing pct for different polling methodologies. The pct distribution varies across methodologies, with some showing greater spread or median differences. This suggests that the polling methodology may influence pct outcomes.

- pct vs days taken from election: A scatter plot displaying pct versus the number of days before the election. The trend line indicates a slight negative relationship, suggesting that as the election date approaches, pct may decrease slightly.

- pct vs state: A boxplot depicting pct across different states. The pct distribution varies by state, with some states showing wider variability or different median values, implying state-specific effects on pct.

```{r}
#| label: fig-pairs
#| echo: false
#| warning: false
# numeric_data <- cleaned_data[sapply(cleaned_data, is.numeric)]
numeric_data = cleaned_data |> select(pct, sample_size, pollscore, days_taken_from_election)

# Create the pairs plot
pairs(numeric_data)
```

@fig-pairs The pairs plot displays scatter plots of four numeric variables (`pct`, `sample_size`, `pollscore`, and `days_taken_from_election`) to visualize their relationships. The data shows clustering, particularly in `pct` versus `sample_size`, suggesting potential heteroscedasticity. The `sample_size` variable is skewed towards lower values, while `pollscore` and `days_taken_from_election` have a more even spread, though `pollscore` shows central clustering. No strong linear relationships are immediately apparent between the variables, indicating that correlations are likely weak.

@@

## Clean Data
@tbl-cleaned-data The data cleaning process involves several steps to ensure the quality and relevance of the polling data. First, we filter the dataset to retain only poll results with a numeric grade of 2.7 or higher, indicating that the polls are considered reliable. Next, we address missing values in the state attribute: polls with NA in the state column are considered national polls.

@@

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Sample of cleaned US election data 
#| echo: false
#| warning: false

cleaned_data |>
  select(pct, sample_size, pollscore, days_taken_from_election, state, methodology, candidate_name) |>
  head(6) |>
  kable(
    col.names = c("pct", "sample_size", "pollscore", "days_taken_from_election","state", "methodology", "candidate_name"),
    booktabs = TRUE
    ) 
```

@@

```{r}
#| label: fig-clean-data-histogram-Harris-Trump
#| fig-cap: the average PCT vs State for Harris and Trump
#| echo: false
#| warning: false
#| fig-subcap: ["average PCT vs State for Harris","PCT vs State for Trump"]
#| layout-nrow: 2
#| fig-width: 40
#| fig-height: 20
Harris_data = cleaned_data[cleaned_data$candidate_name == "Kamala Harris", ]
Harris_data |>
  group_by(state) |>
  summarise(avg_pct = mean(pct, na.rm = TRUE)) |>
  ggplot(mapping = aes(x = state, y = avg_pct)) + 
  geom_col()  +
  theme_minimal() +
  labs(
    x = "State", y = "support percentage(pct)", title = "average support percentage(pct) for Harris per state"
  )

Trump_data = cleaned_data[cleaned_data$candidate_name == "Donald Trump", ]
Trump_data |>
  group_by(state) |>
  summarise(avg_pct = mean(pct, na.rm = TRUE)) |>
  ggplot(mapping = aes(x = state, y = avg_pct)) + 
  geom_col()  +
  theme_minimal() +
  labs(
    x = "State", y = "support percentage(pct)", title = "average support percentage(pct) for Harris per state"
  )

# cleaned_data %>%
#   mutate(month = factor(month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
#                                           "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>%
#   ggplot(mapping = aes(x = month, y = total_decedents)) + 
#   geom_bar(stat = "identity") +
#   theme_minimal() +
#   labs(
#     x = "Month", 
#     y = "Total Decedents"
#   )
```

@fig-clean-data-histogram-Harris-Trump-1 The histogram displays the average support percentage for Kamala Harris across different U.S. states. The data indicates that support for Harris varies significantly across states. Notable observations include relatively high average support in states such as California and New York, which are known for being more Democratic-leaning. On the other hand, there are states with lower average support percentages, particularly in more traditionally Republican or swing states. The distribution suggests regional variations in support, with some states showing consistent backing for Harris while others indicate a weaker performance.

@fig-clean-data-histogram-Harris-Trump-2 The second histogram shows the average support percentage for Donald Trump across various states. It highlights substantial support in states such as Florida and Texas, which align with historical trends of strong Republican support. Trump's average support appears robust in many midwestern and southern states, which are known for their conservative voter base. However, in more liberal-leaning states such as California and New York, the average support is lower, reflecting these states' tendency to lean Democratic.

# Model

The goal of our modeling strategy is twofold. Firstly, we aim to accurately predict the support percentage (PCT) for Harris and Trump based on relevant poll data and key influencing factors. Secondly, we seek to evaluate the efficacy of different modeling approaches—from simple linear regression (SLR) to multiple linear regression (MLR) and Bayesian hierarchical models—to understand their predictive capabilities and assess the underlying relationships between variables. By comparing these models, we can determine which approach provides the most robust and reliable predictions, while considering the variability and potential uncertainty in the data.

@@

The Bayesian model is implemented in R [@citeR] using the rstanarm package as described by @rstanarm. The model is run with the following specifications:

- Formula: $\text{pct} \sim \text{pollscore} + \text{days taken from election} + \text{sample size} + (1 | \text{methodology}) + (1 | \text{state})$

- Priors: Normal(0, 2.5) for all coefficients and intercept, Exponential(1) for $\sigma$

- Settings: Seed = 123, Cores = 4, Adapt delta = 0.95

@@

## Basic Model
```{r}
#| label: fig-model-pct-pollscore
#| echo: false
#| warning: false
just_harris_data = Harris_data |> na.omit()
lm_model1 = lm(pct ~ pollscore, data = just_harris_data)
predictions = predict(lm_model1, just_harris_data)
# summary(lm_model1)

ggplot(just_harris_data, aes(x = pct, y = predictions)) +
  geom_point(color = 'blue') +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Actual vs Predicted Values",
       x = "Actual pct",
       y = "Predicted pct") +
  theme_minimal()
```

@fig-model-pct-pollscore in this SLR model, the response variable is pct and the only one predictor is the pollscore. The primary concern lies in the evident dispersion of data points, which are widely spread and do not cluster closely around the line of perfect prediction (the dashed red line). This suggests that while pollscore may have some predictive capability, it does not adequately explain the variability in pct. The observed inconsistencies between actual and predicted values indicate that the relationship between pct and pollscore is likely not sufficiently captured by a linear model with just one predictor.

```{r}
#| echo: false
#| warning: false
just_harris_data = Harris_data |> na.omit()
model_MLR = lm(pct ~ pollscore + days_taken_from_election + methodology + sample_size + state, data = just_harris_data)
# summary(model_MLR)
just_harris_data <- just_harris_data |> mutate(fitted_value = predict(model_MLR),  num_harris = round((pct / 100) * sample_size, 0))

predictions = predict(model_MLR, just_harris_data)

ggplot(just_harris_data, aes(x = end_date)) +
  geom_point(aes(y = pct), color = "black") +
  geom_line(aes(y = fitted_value), color = "blue", linetype = "dotted") +
  facet_wrap(vars(methodology)) +
  theme_classic() +
  labs(y = "Harris percent", x = "Date", title = "Linear Model: pct ~ end_date + pollster")
```

```{r}
#| label: fig-model-MLR
#| echo: false
#| warning: false
ggplot(just_harris_data, aes(x = pct, y = fitted_value)) +
  geom_point(color = 'blue') +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Actual vs Predicted Values",
       x = "Actual pct",
       y = "Predicted pct") +
  theme_minimal()
```

@fig-model-MLR, in this MLR model, we predicts pct using pollscore, days_taken_from_election, methodology, sample_size, and state as predictors. The overall distribution of points suggests that the MLR model has captured a substantial portion of the variance in pct. The majority of the data points align relatively well with the red dashed line, particularly within the middle range of pct values (approximately between 40 and 60). This alignment indicates that the model performs reasonably well in this range, with predicted values correlating strongly with the actual observed outcomes.

## Bayesian Model
```{r}
#| label: fig-bayesian-model-first
#| echo: false
#| warning: false
bayesian_model_1 = readRDS(here("models/bayesian_model_1.rds"))

pp_check(bayesian_model_1)
# summary(bayesian_model_1)
```

### Posterior Predictive Checks 
@fig-bayesian-model-first This plot displays the results of a posterior predictive check (PPC) for a Bayesian model using the pp_check() function.

The observed data (y) follows a distinct, symmetrical distribution centered around a specific value, suggesting a well-defined peak with tapering on both sides. The replicated distributions (y_rep), shown as multiple thin lines, generally align with the shape of the observed distribution, indicating that the Bayesian model has captured the main characteristics of the data. However, the variability in the y_rep lines highlights the degree of uncertainty inherent in the model’s predictive capability.

The fact that the y_rep curves closely match the overall pattern of the actual y suggests that the model performs reasonably well in replicating the observed data. Minor discrepancies or deviations between the y and y_rep might imply areas where the model could be fine-tuned or adjusted to improve accuracy. Overall, this PPC indicates that the model provides a decent fit to the data, with some variability accounted for in the predictive samples.

### Train Test Validation

```{r}
#| echo: false
#| warning: false
# Split data into training and test sets
set.seed(123)
# train_indices <- sample(seq_len(nrow(just_harris_data)), size = 0.7 * nrow(just_harris_data))
# train_data <- just_harris_data[train_indices, ]
# test_data <- just_harris_data[-train_indices, ]
train_data = read_csv(here("data/02-analysis_data/train_data.csv"))
test_data = read_csv(here("data/02-analysis_data/test_data.csv"))

# Fit the model on the training data
bayesian_model_train = readRDS(here("models/bayesian_model_train.rds"))
# bayesian_model_train <- stan_glmer(
#   formula = formula,
#   data = train_data,
#   family = gaussian(),
#   prior = priors,
#   prior_intercept = priors,
#   seed = 123,
#   cores = 4,
#   adapt_delta = 0.95
# )

# Predict on the test data and check performance
predictions <- posterior_predict(bayesian_model_train, newdata = test_data)
```

```{r}
#| echo: false
#| warning: false
ss_total <- sum((test_data$pct - mean(test_data$pct))^2)
ss_residual <- sum((test_data$pct - colMeans(predictions))^2)
r_squared <- 1 - (ss_residual / ss_total)
print(paste("R-squared:", r_squared))
# "R-squared: 0.700563202695799"
```

\begin{align}
SS_{\text{total}} = \sum_{i=1}^{n} \left( y_i - \bar{y} \right)^2\\
SS_{\text{residual}} = \sum_{i=1}^{n} \left( y_i - \hat{y}_i \right)^2\\
R^2 = 1 - \frac{SS_{\text{residual}}}{SS_{\text{total}}}\\
\text{R-squared} = 0.700563202695799
\end{align}

Where:\
\begin{itemize}
    \item \( y_i \) represents the actual values,
    \item \( \bar{y} \) is the mean of the actual values,
    \item \( \hat{y}_i \) are the predicted values,
    \item \( SS_{\text{total}} \) is the total sum of squares,
    \item \( SS_{\text{residual}} \) is the sum of squared residuals.
\end{itemize}

we could see that the R squared value equal to the 0.700563202695799 suggesting that the model captures a substantial portion of the variability in the data and demonstrates that the model performs well in explaining the variability of the pct in the test data, reflecting strong predictive capabilities.

\newpage
```{r}
#| label: fig-bayesian-model-actual-predict
#| echo: false
#| warning: false
plot(test_data$pct, colMeans(predictions), 
     xlab = "Actual PCT Values", 
     ylab = "Predicted  PCT Values", 
     main = "Predicted PCT vs Actual PCT")
abline(0, 1, col = "red")  # Add a 45-degree reference line
```

 by visual checking actual pct and predicted pct plot @fig-bayesian-model-actual-predict the points are plotted against a 45-degree line, which represents the ideal scenario where predicted values match the actual values perfectly.

\newpage
```{r}
#| label: fig-bayesian-model-residual
#| echo: false
#| warning: false
residuals <- test_data$pct - colMeans(predictions)
plot(colMeans(predictions), residuals, 
     xlab = "Predicted Values", 
     ylab = "Residuals", 
     main = "Residual Plot")
abline(h = 0, col = "red")
```

and the residual plot @fig-bayesian-model-residual are fairly centered around zero with no major trend, suggesting that the model is not heavily biased in its predictions.

In conculsion, the visual checks from the predicted vs. actual plot and the residual plot, there is no strong evidence that the Bayesian model is overfitting. It appears to generalize well to the data it was trained on without showing signs of capturing noise or irrelevant patterns.

@@

To predict the support percentage (PCT) for Harris and Trump, we employed a comprehensive modeling strategy involving Simple Linear Regression (SLR), Multiple Linear Regression (MLR), and a Bayesian hierarchical model. The SLR model provided a foundational analysis of the relationship between PCT and pollscore, highlighting its limited predictive power due to a lack of complexity. The MLR model improved on this by incorporating additional predictors such as days taken from the election, sample size, methodology, and state, enhancing its predictive accuracy and capturing interactions among variables. The Bayesian hierarchical model further refined our approach, incorporating prior knowledge and accounting for variability at group levels (e.g., methodology and state) through random intercepts. This model provided robust uncertainty quantification through credible intervals and demonstrated strong predictive performance with an R-squared value around 0.70, indicating it effectively explained data variability. By leveraging the strengths of these models, particularly the Bayesian approach’s interpretability and robustness, we achieved reliable predictions and a deeper understanding of the factors influencing support percentages.

# Result
```{r}
#| label: fig-bayesian-model-CI
#| echo: false
#| warning: false
# Plot random effects
plot(bayesian_model_1, pars = "(Intercept)", prob = 0.95)
```

@fig-bayesian-model-CI This plot displays the estimated posterior distribution for the intercept parameter in the Bayesian model. The intercept represents the baseline level of support, expressed as a percentage, for Kamala Harris in the U.S. election data used in the analysis. The plot features a point estimate (depicted by the central dot) that indicates the mean or median of the posterior distribution for the intercept, and a horizontal line showing the 95% credible interval, which signifies the range within which the true value of the intercept is likely to fall with 95% probability.

The range of this credible interval spans from approximately 50 to 54 percent, suggesting that the model estimates the baseline level of support for Kamala Harris to be around this range. The relatively narrow width of the interval implies a certain degree of confidence in the estimate, indicating that the data used in the model provided a clear signal for the intercept's value.

In conclusion, the estimated intercept, which represents the baseline percentage of support for Kamala Harris in the U.S. election data analyzed, is centered around 52%, with a credible interval spanning approximately from 50% to 54%. This suggests that, according to the Bayesian model, the underlying support for Kamala Harris.

```{r}
#| label: tbl-bayesian-model-result-summary
#| echo: false
#| warning: false

predictions2 = posterior_predict(bayesian_model_1, newdata = just_harris_data)

predicted_means = colMeans(predictions2)
predicted_intervals = apply(predictions2, 2, quantile, probs = c(0.025, 0.975))

# Combine the results with the actual data
result_summary = data.frame(
  Actual_PCT = just_harris_data$pct,
  Predicted_PCT = predicted_means,
  Lower_CI = predicted_intervals[1, ],
  Upper_CI = predicted_intervals[2, ]
)

kable(head(result_summary, 10), caption = "Head of Bayesian Model Result Summary")

# print(head(result_summary))

```{r}
#| label: fig-bayesian-model-result-CI
#| echo: false
#| warning: false
ggplot(result_summary, aes(x = Actual_PCT, y = Predicted_PCT)) +
  geom_point() +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.2, color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(
    title = "U.S. Election Prediction Results",
    x = "Actual PCT",
    y = "Predicted PCT"
  )
```

@fig-bayesian-model-result-CI The model's predictive performance for Harris's election data appears generally robust, demonstrating a strong alignment between actual and predicted vote percentages. The majority of the data points cluster around the line, indicating reliable predictions, some variance is evident, particularly in the lower and higher actual percentage ranges. This suggests that while the model performs well on average, it may have larger uncertainty in its predictions at the extremes. @tbl-bayesian-model-result-summary The summary table also highlights that most predicted values align closely with actual values, supporting the model’s robustness in capturing the overall trend of election outcomes.

@@
